\textit{MSF: Modulated Sub-graph Finder}

### 2018-04-03 MSF_Paper_FA.tex
Algo and updated tex

To reduce the complexity to score all possible connected sub-graphs
\texttt{MSF} applies a four steps heuristic as described in the
following. The proceeding identification of modulated sub-graphs from
a network by \texttt{MSF} is presented as pseudocode in flowchart in
Fig.~\ref{fig:pseudocodemsf}.

@@

\texttt{MSF} constructs the first sub-graph starting with the genes
associated with the lowest (most significant) \textit{p}-value deduced from
the DGEA. From this seed the sub-graph is attempted to be extended with its
neighboring genes, starting with the most significant one. A single combined
\textit{p}-value is calculated for
the two genes. If the combined \textit{p}-value is smaller than the
minimal individual gene's \textit{p}-value, the extended sub-graph is
accepted. This step is iteratively repeated until no
further extension is accepted. In this case the process starts over
with all remaining genes not yet in a significantly modulated
sub-graph. This step identifies all the trivial sub-graphs that are
modulated in the whole network (Fig.~\ref{fig:algorithm}b).\newline

@@

In the next step initial modulated sub-graphs are used to check if
they could further be extended beyond the immediate neighborhood.
This is done by testing all possible extension paths up to N genes
for all genes in the sub-graph. Again, this step is iteratively
repeated until no further genes are added to the significant
differentially expressed sub-graphs. This steps bridges small gaps of
genes without a clear differential signal in the
DGEA~(Fig.~\ref{fig:algorithm}c).\newline

@@

After detection and extension of the modulated sub-graphs, they are
tested if combined sub-graph score is better than on their own. The
merging of the two sub-graphs is done by depth first search. If the
two sub-graphs merge with the connector of at most N genes (default
two genes) and the combined \textit{p}-value of the merged sub-graph
including the bridging genes in between is less than the individual
\textit{p}-values of the two sub-graphs, the two sub-graphs are merged
together to one big modulated sub-graph~(Fig.~\ref{fig:algorithm}d). This step is
repeated iteratively until no sub-graphs could be merged.\newline

@@

To demonstrate its usefulness \texttt{MSF} is applied to RNAseq data
set of primary human monocyte-derived macrophages (MDMs) infected with
Ebola virus (GSE84188)~\cite{Olejnik}. Ebola Virus (EBOV) belongs to
the Filoviridea family; filamentous, enveloped and single stranded RNA
viruses. EBOV causes hemorrhagic fever in humans, inducing the host
innate and adaptive immune response to be unable to control virus
infection~\cite{Prins}. Until now there are no approved antiviral
drugs for the treatment of Ebola virus infection~\cite{Konde,Rhein}.
The initial targets of EBOV are the macrophages and dendritic immune
cells~\cite{Falasca,Rhein}. EBOV inhibits the critical innate immune
response of the host, which includes the activation of alpha/beta
interferon (IFN-$\alpha / \beta$)~\cite{Prins,Konde,Cardenas}. The
Ebola viral protein VP35 targets the host type~I interferon
IFN-$\alpha / \beta$ to block the early antiviral immune response
~\cite{Prins,Konde,Falasca,Cardenas,Olejnik}. It has been proposed
that IFN-$\alpha / \beta$ should be tested against Ebola for its
antiviral activity through clinical trials~\cite{Konde}. This data was chosen for analysis because it was the most recent EBOV infection data and the infection was carried out in human macrophages, which would help us comprehend the innate immune response of host. The aim of
testing this data with \texttt{MSF} was to identify the
modulated sub-graphs and check if \texttt{MSF} is able to identify the
IFN-$\alpha / \beta$ gene as one of the sources.

@@

 Three main modulated sub-graphs identified by \texttt{MSF} at 6 hpi
 are shown in Fig.~\ref{fig:Sub-graph6hpi}. the figure shows two forms of graphs, one graph has different colored nodes as
 genes associated to different KEGG pathways and how they cross talk. The KEGG
 association and visualization on the modulated sub-graphs was done
 using StringApp~\cite{StringApp} in Cytoscape~\cite{Cyto}. StringApp can retrieve functional enrichment for Gene Ontology terms, KEGG Pathways, and protein domains for a user-specified network. The other graph like a flow chart,
 unveils how more than one stimulus coincides to prevent the
 progression of infection. Sub-graph 1 shows toll-like receptor,
 cytokine, chemokine and jak-stat genes leading to TNF and eventually
 going into apoptosis. Sub-graph 2 shows Extra-cellular matrix (ECM)
 receptor interaction genes which are suggested to be in interaction
 with Ebola glycoprotein (GP)~\cite{Veljkovic}. ECM receptor, Focal
 adhesion and P13k-Akt signaling genes are the sources from one end of
 the sub-graph. P13k-Akt signaling goes to FoxO signaling which
 eventually leads to cell cycle, which at later time-points is
 disrupted. From the other end of sub-graph 2 toll-like recpetor
 combined with cytosolic DNA sensing and cytokine-chemokine signaling
 activates Jak-stat which leads to the start of apoptosis with a
 activation of a few apoptosis genes. Cytokine is also seen to sink
 into natural killer mediated cytotoxicity. Sub-graph 3 shows how some
 of the cytokine and chemokine genes activate the natural killer
 mediated cytotoxicity and apoptosis.

@@

 To assess the robustness and stability of our method, Poisson
 distributed noise was added to the read counts at the 3 time-points. Then DGEA was
 carried out on the disturbed data with the same parameters as for the
 native data using edgeR, followed by analysis with \texttt{MSF}. This procedure was carried out 100 times
 and every time the genes from the sub-graphs identified from noisy
 data were compared to the genes of sub-graphs identified from the
 native data. This was also done for the DEG obtained for each run, a
 cutoff was used to compare the retrieved DEG from noisy data to the
 native data.  The robustness of \texttt{MSF} and the DEG analysis for
 the time-point 6, 24, and 48 hpi are shown in Fig.~\ref{fig:DEGvsMSF}. We were very stringent in
 adding noise to the counts, which could be seen by the DEG analysis
 robustness. For the last time-point (48h~p.i.) 84~\% genes of the
 sub-graphs from native data were retrieved, for 24~h~p.i. robustness was
 about 79~\%. Robustness was 71~\% at 6~h~p.i., this could be due to that the
 infection merely started and the signal transdcution was weak as
 compared to the later time-points.

 To test \texttt{MSF's} correctness simulation was also conducted. For each time-point, one \texttt{MSF} identified sub-graph was selected and  using simulation random graph was generated with the same number of genes as the sub-graphs utilizing the same interaction file. The simulation identified sources were counted. The objective
 was to find less sources in the simulated sub-graphs than observed in the real sub-graphs identified by
 \texttt{MSF}. Simulation results showed a chance of $<$10$\%$
 of the 1000 random sub-graphs generated to get more sources than observed.

@@

To further understand the functional role of the sub-graphs identified
by \texttt{MSF}, we linked the sub-graphs to the predefined KEGG
pathways and Reactome pathways through gene enrichment analysis. We
used Cytoscape~\cite{Cyto} plugins to identify the enriched patwhays,
CytoKegg~\cite{Cytokegg} for KEGG and Reactome FI~\cite{Reactome} for
Reactome. A comparison to other tools was performed on the same dataset using each software's default parameters. We compared the pathways enriched in \texttt{MSF} identified
sub-graphs to pathway analysis by SPIA~\cite{Tarca} for KEGG and
Reactome pathway enrichment analysis for Reactome.

@@

Likewise we compared gene enrichment results of sub-graph genes
identified by \texttt{MSF} using Reactome FI to the Reactome
predefined pathway enrichment analysis. Although \texttt{MSF} and
Reactome showed good agreement on the dis-regulated pathways e.g
Signaling by Interleukins, Interleukin-10 signaling, Cytokine
Signaling in Immune system, Chemokine receptors bind chemokines,
RIG-I/MDA5 mediated induction of IFN-alpha/beta pathways, NF-kB
activation, Reactome as well could not identify Toll-like receptor
signaling. This shows that
\texttt{MSF} is able to identify the important modulated genes in the
sub-graphs at the earliest time-point of infection.

@@

The discussion should include the implications of the article results
in view of prior work in this field.

@@

\begin{figure*}
	\centering
	\fbox{
		\begin{minipage}{17 cm}
		\includegraphics[width=1.0\linewidth]{PseudoCodeMSF}
		\caption{Shows the algorithm of MSF with four steps to find the modulated sub-graphs and identify sources and sinks.}
		\label{fig:pseudocodemsf}
	\end{minipage}
}

>>>>

To reduce the complexity to score all possible connected sub-graphs
\texttt{MSF} applies a four steps heuristic as described in the
following. The proceeding identification of modulated sub-graphs from
a network by \texttt{MSF} is presented as a graphical pseudo code  in
Fig.~\ref{fig:pseudocodemsf}.

@@

\texttt{MSF} constructs the first sub-graph starting with the genes
associated with the lowest (most significant) \textit{p}-value deduced from
the DGEA. From this seed the sub-graph is attempted to be extended with its
neighboring genes, starting with the most significant one. A single combined
\textit{p}-value is calculated for
the two genes. If the combined \textit{p}-value is smaller than the
minimal individual gene's \textit{p}-value, the extended sub-graph is
accepted. This step is iteratively repeated until no
further extension is accepted. In this case the process starts over
with all remaining genes not yet in a significantly modulated
sub-graph. This step identifies all the trivial sub-graphs that are
modulated in the whole network.\newline

@@

In the next step initial modulated sub-graphs are used to check if
they could further be extended beyond the immediate neighborhood. This is done by testing all possible extension paths up to N genes
for all genes in the sub-graph. Again, this step is iteratively
repeated until no further genes are added to the significant
differentially expressed sub-graphs. This steps bridges small gaps of
genes without a clear differential signal in the
DGEA.\newline

@@

After detection and extension of the modulated sub-graphs, they are
tested if combined sub-graph score is better than on their own. The
merging of the two sub-graphs is done by depth first search. If the
two sub-graphs merge with the connector of at most N genes (default
two genes) and the combined \textit{p}-value of the merged sub-graph
including the bridging genes in between is less than the individual
\textit{p}-values of the two sub-graphs, the two sub-graphs are merged
together to one big modulated sub-graph. This step is
repeated iteratively until no sub-graphs could be merged.\newline

@@

To demonstrate its usefulness \texttt{MSF} is applied to RNAseq data
set of primary human monocyte-derived macrophages (MDMs) infected with
Ebola virus (GSE84188)~\cite{Olejnik}. Ebola Virus (EBOV) belongs to
the Filoviridea family; filamentous, enveloped and single stranded RNA
viruses. EBOV causes hemorrhagic fever in humans, inducing the host
innate and adaptive immune response to be unable to control virus
infection~\cite{Prins}. Until now there are no approved antiviral
drugs for the treatment of Ebola virus infection~\cite{Konde,Rhein}.
The initial targets of EBOV are the macrophages and dendritic immune
cells~\cite{Falasca,Rhein}. EBOV inhibits the critical innate immune
response of the host, which includes the activation of alpha/beta
interferon (IFN-$\alpha / \beta$)~\cite{Prins,Konde,Cardenas}. It has been proposed
that IFN-$\alpha / \beta$ should be tested against Ebola for its
antiviral activity through clinical trials~\cite{Konde}. Since the data was most recent and the  infection was carried out in human macrophages, it helped us comprehend the innate immune response of host. The aim of
testing this data with \texttt{MSF} was to identify the
modulated sub-graphs and check if \texttt{MSF} is able to identify the
IFN-$\alpha / \beta$ gene as one of the sources.

@@

 Three main modulated sub-graphs identified by \texttt{MSF} at 6 hpi
 are shown in Fig.~\ref{fig:Sub-graph6hpi}. It shows two forms of graphs, one graph has different colored nodes as
 genes associated to different KEGG pathways. The KEGG
 association and visualization on the modulated sub-graphs was done
 using StringApp~\cite{StringApp} in Cytoscape~\cite{Cyto}. StringApp can retrieve functional enrichment for Gene Ontology terms, KEGG Pathways, and protein domains for a user-specified network. The other graph is a flow chart, that
 unveils how more than one stimulus coincides to prevent the
 progression of infection. Sub-graph 1 shows toll-like receptor,
 cytokine, chemokine and jak-stat genes leading to TNF and eventually
 going into apoptosis. From one end  of sub-graph 2, Extra-cellular matrix (ECM)
 receptor interaction genes are modulated, which are suggested to be in interaction
 with Ebola glycoprotein (GP)~\cite{Veljkovic}. ECM receptor, Focal
 adhesion and P13k-Akt signaling genes are the sources from this end of
 the sub-graph, with P13k-Akt signaling going to FoxO signaling which
 eventually leads to cell cycle, that at later time-points is
 seen to be disrupted. From the other end of sub-graph 2 toll-like recpetor
 combined with cytosolic DNA sensing and cytokine-chemokine signaling
 activates Jak-stat which leads to the start of apoptosis with a
 few apoptosis genes being modulated. Cytokine is also seen to sink
 into natural killer mediated cytotoxicity. Sub-graph 3 shows how some
 of the cytokine and chemokine genes activate the natural killer
 mediated cytotoxicity and apoptosis.

@@

 To assess the robustness and stability of our method, Poisson
 distributed noise was added to the read counts of the three time-points. Then DGEA was
 carried out on the disturbed data with the same parameters as for the
 native data using edgeR, followed by analysis with \texttt{MSF}. This procedure was carried out 100 times
 and every time the genes from the sub-graphs identified from noisy
 data were compared to the genes of sub-graphs identified from the
 native data. This was also done for the DEG obtained for each run, a
 cutoff was used to compare the retrieved DEG from noisy data to the
 native data.  The robustness of \texttt{MSF} and the DEG analysis for
 the time-point 6, 24, and 48 hpi are shown in Fig.~\ref{fig:DEGvsMSF}. We were very stringent in
 adding noise to the counts, which could be seen by the DEG analysis
 robustness. For the last time-point (48h~p.i.) 84~\% genes of the
 sub-graphs from native data were retrieved, for 24~h~p.i. recall rate was
 79~\%. Robustness was 71~\% at 6~h~p.i., this could be due to that the
 infection merely started and the signal transdcution was weak as
 compared to the later time-points.

 To test \texttt{MSF's} correctness simulation was also conducted. For each time-point, one \texttt{MSF} identified sub-graph was selected and using simulation random graph was generated with the same number of genes as the sub-graphs utilizing the same interaction file. The simulation identified sources were counted from the simulated sub-graph. The objective
 was to find less sources in the simulated sub-graphs than observed in the real sub-graphs identified by
 \texttt{MSF}. Simulation results showed a chance of $<$10$\%$
 of the 1000 random sub-graphs generated to get more sources than observed.

@@

 A comparison to other tools was performed on the same dataset using each software's default parameters. We compared the gene enrichment analysis in \texttt{MSF} identified
sub-graphs to pathway analysis by SPIA~\cite{Tarca} for KEGG and
Reactome pathway enrichment analysis for Reactome.  We
used Cytoscape~\cite{Cyto} plugins to identify the enriched patwhays,
CytoKegg~\cite{Cytokegg} for KEGG and Reactome FI~\cite{Reactome} for
Reactome for \texttt{MSF} identified
sub-graphs.

@@

Likewise we compared gene enrichment results of sub-graph genes
identified by \texttt{MSF} using Reactome FI to the Reactome
predefined pathway enrichment analysis. Although \texttt{MSF} and
Reactome showed good agreement on the dis-regulated pathways e.g
Signaling by Interleukins, Interleukin-10 signaling, Cytokine
Signaling in Immune system, Chemokine receptors bind chemokines,
RIG-I/MDA5 mediated induction of IFN-alpha/beta pathways, NF-kB
activation, Reactome as well could not identify Toll-like receptor
signaling. This shows that
\texttt{MSF} is able to identify the important modulated genes of the early infection stage.

@@

MSF is able to identify the connected modulated genes, showing how the signal transmission flows between the different predefined pathways. MSF is able to identify the important genes at the early stage of infection which at the later stage of infection became the most modulated genes. Even with the addition of noise MSF identifies the important modulated core set of genes. The sources identified by MSF can assist in the discovery of new drug targets. One limitation for MSF would be an incomplete interaction network. For the unknown or less known genes, the interactions are still not well known and so are usually missing in the biological network in different databases.

@@

\begin{figure*}
	\centering
	\fbox{
		\begin{minipage}{17 cm}
		\includegraphics[width=1.0\linewidth]{AlgorithmMSF.png}
		\caption{Shows the algorithm of MSF with four steps to find the modulated sub-graphs and identify sources and sinks.}
		\label{fig:pseudocodemsf}
	\end{minipage}
}
